### 2013-05-17 context-dependence.tex
more writing

Using this observation, we can estimate parameters by composite likelihood,
using all $N(x,y)$ as data.

To get confidence intervals, we need to either simulate, or do something different.
If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
(using Stein's method)
Then we'll have the correct likelihood function for (this subset of) data, and can go Bayesian.

@@

@@

>>>>

Using this observation, we can estimate parameters by composite likelihood,
using all $N(x,y)$ as data.
Maximum likelihood then gets an asymptotically unbiased estimate of the parameters.

To get confidence intervals, we need to either simulate, or do something different.

If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
(using Stein's method)
Then we'll have the correct likelihood function for (this subset of) data.
If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
and any overlapping patterns must require additional changes, 
i.e.\ if $(x_1,y_1)$ matches at $i_1$, 
then for $(x_2,y_2)$ to match at $i_2$, 
there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.

@@

%%%%%%%%
\section{Phylogenetic inference}

In phylogenetic applications, rather than ``before'' and ``after'' observations,
we get two (or more) observations evolved from a common root.
In the simplest case of two taxa we have two processes $X$ and $Y$,
with identical starting states $X(0)=Y(0)$,
observed only at times $t_X$ and $t_Y$ respectively.

If the approximation described above holds,
then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
using the pattern frequencies at the root.
Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
% \[
%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
% \]
The probability that we see $x$ and $y$ at the random location $I$ is
\begin{align} \label{eqn:phylo_likelihood}
    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
\end{align}
This can be computed without much more effort than the simpler case above,
but requires the frequencies at the root.

\subsection{The root distribution}

We clearly don't want to add an extra parameter $\pi(z)$ for all $z \in \calS^{m+2\ell}$. 
If nothing else, it would make the inference depend on $m$ and $\ell$, which we'd like to avoid.
One approach would be to put a Dirichlet distribution on $\pi(z)$,
but this would destroy many of the nice computational properties of the likelihoods,
since making the frequencies of patterns at the root random introduces correlations between counts.
An alternative is to specify $\pi$ through lower-dimensional statistics,
for instance to model the root state as a sample from a Markov random field
with parameters for the marginal frequencies of each state, and pairwise (or higher) correlations between them.

A Markov random field with interactions between sites no more than distance two away 
is determined by the energy function $\phi(a,b,c)$ which gives the ``energy'' of the configuration $abc$ for each $a,b,c \in \calS$.
We use this to define the distribution on root patterns $\pi$ by
\begin{align}  \label{eqn:root_mrf}
    \pi(z) = \frac{1}{Z} \exp\left( - \sum_{i=1}^{|z|} h(z_{i-1},z_i,z_{i+1}) \right),
\end{align}
i.e.\ the exponential of the total energy of each three adjacent sites,
where $Z$ is the normalizing constant.
Note that it is often difficult to compute $Z$ in applications of Markov random fields,
but here we are working with relatively short sequences, 
so we just compute \eqref{eqn:root_mrf} for all possible $z$, and divide by the sum.
(However, for this same reason this is not equivalent to saying that $X(0)$ is a sample from the Markov random field
and finding the distribution of patterns within this.)

@@

For inference, we need to compute the $|\calS|^{m+2\ell} \times |\calS|^m$ matrix whose $(x,y)^\text{th}$ entry is $p_{m,\ell}(t,x,y)$,
for various $t$.
This matrix is a projection of the $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix whose $(x,z)^\text{th}$ entry is
\begin{align}
    p_{m+2\ell}(x,z) = \left( e^{t G(m+2\ell)} \right)_{x,z} ,
\end{align}
where $G(m+2\ell)$ is the sparse $|\calS|^{m+2\ell} \times |\calS|^{m+2\ell}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long patterns $z$ that match the shorter pattern $y$:
if we define the matrix $U$ in this way, so that $U_{zy}=1$ if $z_\ell^m=y$ and $U_{zy}=0$ otherwise,
then
\[
    p_{m,\ell}(t,x,y) = \sum_{z \in \calS^{m+2\ell}} \left( e^{t G(m+2\ell)} \right)_{xz} U_{zy} ,
\]
so in fact we don't need the entire matrix $e^{t G(m+2\ell)}$,
just the product of this matrix with each of the columns of $U$.

Modern techniques in sparse matrix computation (e.g.\ Krylov methods) provide efficient ways to do this
by only multiplying $G$ and $U$ some number of times, 
which is much more efficient because $G$ is sparse.

Also note that the further calculations in the phylogenetic case can be done using these methods.
To compute
\[
    \sum_z \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) 
    = \sum_z \left( e^{t_X G(m+2\ell)} \right)_{zx} \pi(z) \sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy} ,
\]
we can first compute $\sum_w \left( e^{t_X G(m+2\ell)} \right)_{zw} U_{wy}$ as above,
multiply by $\pi(z)$, and then apply the algorithm again using the transpose of $G(m+2\ell)$.

Using the R package expm, this makes computation quite feasible:
with four possible states and $\ell=2$ and $m=1$, so that $G(m+2\ell)$ is a $1024 \times 1024$ matrix,
computing $e^{t G(5) }$ takes 13 seconds, while computing $e^{t G(5)} U$ takes only 0.3 seconds.
Increasing to $\ell=4$ and $m=1$ or $\ell=3$ and $m=2$ is still feasible, taking $e^{t G(8)} U$ in 42 and 47 seconds, respectively.
